On the uniqueness and stability of dictionaries for sparse representation of noisy signals

### 2017-04-01 manuscripts/PNAS/chazthm_pnas.tex
discussion SR

In this note, we generalize the approach of \cite{Hillar15} for showing uniqueness in Prob.~\ref{InverseProblem} to the case of noisy measurements while also reducing the sufficient number of samples.
% FRITZ: Inverse problems instead of Data Analysis. Don't assume that everyone is assuming they are recovering ground truth. Cite results again and state implications! Focus on them (probabilistic ones too).
Our results provide theoretical grounding for the use of sparse linear coding in blind source separation, wherein the goal is to infer the generating dictionary and sparse codes from noisy measurements generated as in \eqref{LinearModel}.  Such results are motivated by the seemingly universal representations discovered by training a sparse linear coding model with natural data, independently of specific algorithmic implementation.  We also collect a set of useful mathematical tools and basic results for future research.  % We close with several application areas.

%\textbf{Inverse Problems}.  
% from $N=k{m \choose k}^2$ to $N = m(k-1){m \choose k}+m$. 
% CUT OK? Surprisingly, almost all $n \times m$ dictionaries satisfying the standard assumption \eqref{CScondition} from compressed sensing (CS) are identifiable from enough generic noisy $k$-sparse linear combinations of their elements, up to an error linear in the noise. Moreover, if solutions are constrained to satisfy \eqref{SparkCondition}, then only an upper bound on the number of dictionary elements need be taken as given. 
One interesting aspect of our contribution is a combinatorial criteria (regular hypergraphs satisfying the singular intersection property, Def.~\ref{sip}) for choosing support sets for original sparse codes. Fully understanding those combinatorial designs allowing for stable representations of datasets is an interesting research area for the future.
We also remark that our main uniqueness result (Thm.~\ref{DeterministicUniquenessTheorem}) accounts for deterministic ``worst-case'' noise, whereas the ``effective'' noise might be much smaller when it is sampled from a given distribution; in such cases, the constants in our theorems %Thms.~\ref{DeterministicUniquenessTheorem},~\ref{DeterministicUniquenessTheorem2} 
should be better. %We note also that these results extend trivially to cases where point-wise injective nonlinearities are applied to the data.

As an inverse problem to be approximated by a (typically) numerical algorithm, what we have shown here is that the sparse linear coding model  generally produces \textit{well-posed} optimizations.  This early concept of Hadamard \cite{Hadamard1902}) can be paraphrased as the idea that algorithms inverting a parameterized model of data should have unique solutions that are continuous in the data.  In other words, perturbing slightly a dataset should only slightly influence the uniqueness of parameters in the model, rather than discontinuously affect the character of optimization involved.  In this regard, we provide explicit constants in terms of original latent parameters, although they are too expensive to compute for any practical utility.    % We hope that researchers in sparse linear coding continue to improve on the bounds and constants here.

One of the technical difficulties in our main theorem is the absence of a spark condition assumption on alternate dictionaries; in fact, several results here are much simpler to prove when such an assumption is made.  However, in general, it cannot necessarily be assumed that a learned dictionary have satisfy the spark condition.  In addition, guaranteeing that a dictionary has this property is probably intractable given its NP-hardness \cite{tillmann2014computational}.  It would be of practical utility to determine the best possible dependence of $\varepsilon$ on $\delta_1, \delta_2$ (see \eqref{epsdel}) as well as the minimal requirements on the number and diversity of generating codes. %We encourage researchers to extend our results and tighten these parameters.

%\textbf{Theoretical Neuroscience}.
In the field of theoretical neuroscience, sparse dictionary learning and related methods have recovered characteristic components of natural images and sounds \cite{bellsejnowski1996, smithlewicki2006, Carlson12}, reproducing response properties of cortical neurons. Our theorems suggest that this correspondence could be due to the uniqueness of sparse representations. Furthermore, our guarantees justify the hypothesis of \cite{Coulter10, Isely10} that sparse codes passed through a communication bottleneck in the brain can be recovered from random projections via (unsupervised) biologically plausible sparse dictionary learning (e.g., \cite{rehnsommer2007, rozell2007neurally, hu2014hebbian}).  In fact, the concept of ``universal'' objects in the brain is an early idea in neuroscience theory \cite{pitts1947}. But this is also in data science applications.  For instance, when recovering a rat's position on a linear track from learning a sparse linear model on  local field potentials in Hippocampus \cite{Agarwal14}.

@@

%\textbf{Engineering}.
In engineering, several groups have utilized compressed sensing for signal processing tasks: MRI analysis \cite{lustig2008compressed},  image compression \cite{Duarte08}, and, more recently, the design of an ultrafast camera \cite{Gao14}. Given such effective uses of CS, it is only a matter of time before these systems incorporate sparse dictionary learning to encode and process data. Guarantees such as those offered by our theorems allow any such device to be equivalent to any other (having different initial parameters and data samples) as long as enough data originate from a statistically identical system.

>>>>

In this note, we generalize the approach of \cite{Hillar15} to prove uniqueness in Prob.~\ref{InverseProblem} for the case of noisy measurements while also reducing the sufficient number of samples.
% FRITZ: Inverse problems instead of Data Analysis. Don't assume that everyone is assuming they are recovering ground truth. Cite results again and state implications! Focus on them (probabilistic ones too).
Our results provide theoretical grounding for the use of sparse linear coding in blind source separation, wherein the goal is to infer the generating dictionary and sparse codes from noisy measurements, \eqref{LinearModel}.  We also collect a set of useful mathematical tools and basic facts for future research.
The main motivation for this work, however, was to understand how seemingly universal and stable representations are discovered by training a sparse linear coding model on natural data, independent of specific algorithmic implementation.    We elaborate more on these applications below after discussing  theoretical aspects.  % We close with several application areas.

%\textbf{Inverse Problems}.  
% from $N=k{m \choose k}^2$ to $N = m(k-1){m \choose k}+m$. 
% CUT OK? Surprisingly, almost all $n \times m$ dictionaries satisfying the standard assumption \eqref{CScondition} from compressed sensing (CS) are identifiable from enough generic noisy $k$-sparse linear combinations of their elements, up to an error linear in the noise. Moreover, if solutions are constrained to satisfy \eqref{SparkCondition}, then only an upper bound on the number of dictionary elements need be taken as given.

What we have shown here is that as an inverse problem to be approximated by a numerical algorithm, the sparse linear coding model  generally produces \textit{well-posed} optimizations.  This early concept of Hadamard \cite{Hadamard1902} can be paraphrased as the idea that good parameterized models should pose optimization problems with solutions that are stable; that is, continuous in the data.  In other words, perturbing a dataset by a small amount should only slightly influence the identifiability or uniqueness of parameters in the model, rather than discontinuously affect the character of optimization involved.  In this regard, we demonstrate a linear relationship between noise in the data and the unique solution in  sparse linear coding, with explicit constants expressed in terms of original latent parameters. 
Nonetheless, it would be of practical utility to determine the best possible dependence of $\varepsilon$ on $\delta_1, \delta_2$ in \eqref{epsdel} as well as the minimal requirements on the number and diversity of generating codes, and we hope that other researchers continue to improve and extend our results.
% , although they are too computationally expensive to be of practical value.  We hope .

One notable component of our contribution is a combinatorial criteria (regular hypergraphs satisfying the singular intersection property, Def.~\ref{sip}) for choosing support sets for original sparse codes. Fully understanding those combinatorial designs allowing for stable representations of datasets is an interesting research area for the future.
We also remark that our main uniqueness result (Thm.~\ref{DeterministicUniquenessTheorem}) accounts for deterministic ``worst-case'' noise, whereas the ``effective'' noise might be much smaller when it is sampled from a given distribution; in such cases, the constants in our results %Thms.~\ref{DeterministicUniquenessTheorem},~\ref{DeterministicUniquenessTheorem2} 
will improve. %We note also that these results extend trivially to cases where point-wise injective nonlinearities are applied to the data.

A technical difficulty in our main theorem is the absence of a spark condition assumption on alternate data factorizations.
Although mathematically interesting that no such requirement is necessary, there are other reasons to desire such a theoretical guarantee. For instance, it is difficult to ensure an algorithm maintain a dictionary satisfying \eqref{SparkCondition} at each learning stage; indeed, even certifying that a dictionary has this property is probably intractable given its NP-hardness \cite{tillmann2014computational}.

In fact, though, uniqueness guarantees with minimal assumptions apply to all areas of data science and engineering that utilize learned sparse representations.  For example, several groups have utilized compressed sensing for signal processing tasks: MRI analysis \cite{lustig2008compressed},  image compression \cite{Duarte08}, and, more recently, the design of an ultrafast camera \cite{Gao14}. Given such effective uses of CS, it is only a matter of time before these systems incorporate sparse representational learning to encode and process data (e.g., in a device that learns structure from motion \cite{kong2016prior}).  In these cases, guarantees such as those offered by our theorems allow any device to be equivalent to any other (having different initial parameters, data samples, and perhaps algorithms) as long as enough data originate from a statistically identical system.
%Data science:
%  
%in a common application of compressed sensing in engineering

%\textbf{Theoretical Neuroscience}.
Additionally, in the field of theoretical neuroscience, sparse dictionary learning and related methods have recovered characteristic components of natural images \cite{Olshausen96, hurri1996image, bell1997independent, van1998independent} and sounds \cite{bellsejnowski1996, smithlewicki2006, Carlson12}, reproducing response properties of cortical neurons. Our theorems suggest that this correspondence could be due to  sparse representations being ``universal'' in natural data, an early mathematical idea in neural theory \cite{pitts1947}. Furthermore, our guarantees justify the hypothesis of \cite{Coulter10, Isely10} that sparse codes passed through a communication bottleneck in the brain can be recovered from random projections via (unsupervised) biologically plausible sparse dictionary learning (e.g., \cite{rehnsommer2007, rozell2007neurally, hu2014hebbian}).

%In fact, the concept of ``universal'' objects in the brain is .

%But this is also in data science applications.  For instance, when recovering a rat's position on a linear track from learning a sparse linear model on  local field potentials in Hippocampus \cite{Agarwal14}.

@@

%\textbf{Engineering}.
